Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-07 submission/manuscript.Rmd
cleaning up intro, need to add germline bacgkround

Embryonic stem cells employ a single genome to form the myriad of discrete cellular identities that make up a complete organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through histone and DNA modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in cellular identity [@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by or linked to mutations in chromatin regulators[CIT]. The connection between neurodevelopment and chromatin regulation can be partially explained by their regulation of brain-specific genes or chromatin states, such as orchestrating synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] or the transition between neuron and glia during cortical development[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in aberrant transcription of tissue-specific genes outside of their intended environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have explored the ectopic expression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate if dysregulation of brain tissue identity contributes to neurodevelopmental impairments, it is essential to first characterize the types of genes misexpressed and the molecular mechanisms underlying their de-repression. In this study, we characterized the dysregulation of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been characterized. Furthermore, some of the ectopic testis transcripts have known functions in germ cells[@scandagliaLossKdm5cCauses2017], potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss. While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
- KDM5C's molecular mechanisms unclear - direct vs indirect
- Developmental time point of involvement
- How the genes change over time.

@@

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, ovary, and testis genes. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are typically only expressed in germ cells. While the _Kdm5c_-KO brain primarily expresses genes important for late germ cell development, _Kdm5c_-KO EpiLCs aberrantly express key drivers of germline identity and meiosis, such as _Dazl_ and _Stra8_. We found that KDM5C only binds a subset of germline genes expressed in the _Kdm5c_-KO brain and EpiLCs, indicating germline-enriched mRNAs can be aberrantly expressed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

- We then generated a dataset of germline-enriched genes to aid in the characterization of germline gene misexpression
- **note: need a better conclusion sentence - work on when we know what's happening with last figure/functional consequences**
- **note: I feel like I'm missing the key points still on the last paragraph**

>>>>

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Although many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a], recent advancements in next generation sequencing unexpectedly revealed their dysregulation underlies many neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. Many studies have suggested this connection between neurodevelopment and chromatin regulators is due to their regulation of brain-specific genes, such as orchestrating the transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and the transition between neuronal and glial differentiation in neural precursor cells[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Very few studies have investigated the misexpression of tissue-specific genes in chromatin-linked neurodevelopmental disorders[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unclear if this partial loss of brain identity contributes to neurodevelopmental impairments.

To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the types of genes dysregulated and the molecular mechanisms underlying their de-repression. In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C). KDM5C is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO brain unexpectedly revealed ectopic expression of testis genes within the hippocampus[@scandagliaLossKdm5cCauses2017]. It is currently unclear if these testis genes impair _Kdm5c_-KO neurodevelopment and if this dysregulation in _Kdm5c_-KO brain tissue identity is unique to testis genes.

Intriguingly, some of the ectopic testis transcripts expressed within the _Kdm5c_-KO brain have known functions in germ cells[@scandagliaLossKdm5cCauses2017], suggesting KDM5C may play a role in demarcating somatic and germline identity.

- germ line background

potentially reflecting a failure to distinguish between somatic (non-meiotic) and germline (meiotic) identities with KDM5C loss. While there is emerging evidence that KDM5C regulates select germline genes, it is unclear which germline  _Kdm5c_-KO.
- KDM5C's molecular mechanisms unclear - direct vs indirect
- Developmental time point of involvement
- How the genes change over time.

@@

- an _in vitro_ model of the post-implantation embryo.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the male _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs). We observed general dysregulation of tissue-enriched genes the _Kdm5c_-KO brain and EpiLCs, including misexpression of genes typically highly enriched within the testis, liver, muscle, and ovary. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are typically unique to germ cells. While the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis, _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_. KDM5C was highly enriched at germline gene promoters in EpiLCs but only bound to a subset of germline genes expressed in _Kdm5c_-KO cells, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the embryonic silencing of germline genes in somatic cells by aiding the placement of DNA methylation in EpiLCs through H3K4me3 removal. Thus, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

- **note: last sentence needs work**
